neonstore: provenance-aware persistent storage for NEON data

### 2021-12-21 paper/paper.Rmd
:pen:
support lazy-table

The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing envionmental science [@NAS_NEON; @NAS_GRAND].

@@

Provenance and performance are two central aspects to the design of `neonstore`.  A focus on provenance means that analyses can always be traced back to individual NEON data files, and supports both workflows that require access to the most recent or updated data and workflows that require repeated access to persistent, stable data files.  A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available RAM on most machines (e.g. 10 GB - 10 TB range, depending on products and sites considered).

@@

RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #].

@@

```{r}
library(neonstore)
```

@@

The workhorse function of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.  For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`: (The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling).

@@

@@

`neon_index()` takes optional arguments such as `product`, `start_date`, `end_date`, and `site` to allow for convenient sub-setting.  Because most observational and instrument-based products are divided by site and month, this is still not a convenient way to browse the data.

```{r}
neon_read("brd_countdata-expanded")
```

Two other functions access additional API endpoints 
that may also be of interest. `neon_sites()` returns a `data.frame`
of site information, including site descriptions and 
the ecological domain that each site falls into:

```{r}
neon_sites()
```

Lastly, `neon_products()` returns a table with a list of all neon products,
which may be useful for data discovery or additional metadata about any
given product:

```{r}
neon_products()
```

Note that at this time, `neonstore` does not provide a complete

## Design Details / comparison to `neonUtilities`

`neonstore` is not meant as a replacement to the `neonUtilities` package
developed by NEON staff.  `neonUtilities` performs a range of product-specific
data querying, parsing, and data manipulation beyond what is provided by NEON's
API or web interface. `neonUtilities` also provides other utilities for working 
with NEON data beyond the scope of the NEON API or the data download/ingest 
process. While this processing is undoubtedly useful, it may make
it difficult to compare results or analyses based on data downloaded and accessed
using `neonUtilities` R package with analyses based on data accessed directly
from the web interface, the API, or another tool (or even a different release
of the `neonUtilities`).

By contrast, `neonstore` aims to do far less.  `neonstore` merely automates the
download of individual NEON data files.  In contrast to `neonUtilities` which 
by default "stacks" these raw files into single tables and discards the raw 
data, `neonstore` preserves only the raw files in the store, stacking the 
individual tables "on demand" using `neon_read()`. `neon_read()` is a thin
wrapper around the `vroom` package, 
[Hester & Wickham, 2020](https://vroom.r-lib.org), which uses the `altrep` 
mechanism in R to provide very fast reads of rectangular text data into R,
and trivially handles the case of a single table being broken across many files.
Some NEON tables are not entirely consistent in their use of columns across the
individual site-month files, so `neon_read()` transparently checks for this,
reading in groups of files sharing all matching columns with `vroom` before
binding the groups together. This makes it easier to always trace an analysis
back to the original input data, makes it easier to update input data files 
without facing the challenge of either downloading & stacking the whole 
data product from scratch again or having to keep track of some previously
downloaded data file.

A few other differences are also worth noting.

- `neonstore` aims to provide persistent storage, writing raw data files to
  the appropriate app directory for your operating system (see `rappdirs`,
  [Ratnakumar et al 2016](https://CRAN.R-project.org/package=rappdirs)). 
  More details about this can be found in [Provenance](/#Provenance), below.
- `neon_download()` provides  clean and concise progress bars for the two key
  processes involved: querying the API to obtain download URLs (which involves no 
  large data transfer but counts against API rate limiting, see below), and the
  actual file downloads.
- `neon_download()` will verify the integrity of file downloads against the MD5
  hashes provided. 
- `neon_download()` will omit downloads of any existing data files in the local 
  store.  
- You can request multiple products at once using vector notation, though API
  rate limiting may interfere with large requests.
- `neon_download()` uses `curl::curl_download()` instead of `downloadr` package
  used in `neonUtilities`, which can be finicky on Windows and older versions of R. 
- `neonstore` has slightly lighter dependencies: only `vroom` and `httr`, and
  packages already used by one of those two (`curl`, `openssl`).

Like `neonUtilities`, You can optionally include site and date filters,
e.g. to request only records more  recent than a certain date.  Doing so will
preserve API quota and improve speed (see API limits, below). `neonUtilities`
is also far more widely tested and has extensive error handling tailored to
individual data products.

Because `neonstore` only stores raw data products as returned from the NEON API,
it can easily determine which files have already been downloaded, and only
download new files without requiring the user to specify specific dates. 
(It must still query the API for all the metadata in the requested date range).
This same modular approach also makes it easy to track _data provenance_, an 
essential element of reproduciblity in comparing results across other analyses
of the NEON data.

We can list precisely which component files are being read in by `neon_read()`
by consulting `neon_index()`:

```{r}
raw_files <- neon_index(table = "brd_countdata-expanded", hash="md5")
raw_files
```

`neon_read()` is a relatively trivial function that simply passes this file 
list to `vroom::vroom()`, a fast, vectorized parser that can easily read in 
a single table that is broken into many separate files.

Imagine instead that we use the common pattern of downloading
these raw files, stacks and possibly cleans the data, saving only this derived
product while discarding the individual files.  Now imagine a second researcher,
at some later date, queries the API over the same reported range of dates and 
sites, uses the same software package to stack the tables, only to discover the
resulting table is somehow different from ours (e.g. by comparing file hashes).
Pinpointing the source of the discrepancy would be challenging and 
labor-intensive.

In contrast, the same detective-work would be easy with the `neonstore` file
list.  We can confirm if the API had returned the same number of 
raw files with the same names; and better, can verify integrity of the contents
by comparing hashes of files now being returned to those recorded
by `neon_index()`.  In this way, we could determine if any additional files
had been included or pinpoint any files that may have changed.

As such, users might want to store the `neon_index()` `data.frame` for the
table(s) they have used as part of their analysis, including the individual
file hashes.  One can also generate a zip of all the data files for 
archival purposes. (Note that NEON is an Open Data provider, see
[LICENCE](https://www.neonscience.org/data/about-data/data-policies).)

```{r message=FALSE}
write.csv(raw_files, "index.csv")
neon_export("brd_countdata.zip")
```

## Data citation

Always remember to cite your data sources!
`neonstore` knows how to generate the appropriate citation for the data
in your local store (or any specific product).

```{r}
## Note: NEON's recommendation may soon change
neon_citation()
```

## Note on API limits

[The NEON API now rate-limits requests.](https://data.neonscience.org/data-api/rate-limiting/#api-tokens).
Using a personal token will increase the number of requests you can make.
See that link for directions on registering for a token.
Then pass this token in `.token` argument of `neon_download()`,
or for frequent use, add this token as  an environmental variable, `NEON_DATA`
to your local `.Renviron` file in your user's home directory.

`neon_download()` must first query each the API of eacn NEON site which collects
that product, for each month the product is collected.
(It would be much more efficient on the NEON server if the API could take
queries of the from `/data/<product>/<site>`, and pool the results, rather than
require each month of sampling separately!)

```{r include=FALSE}
unlink("brd_countdata.zip")
unlink("index.csv")
Sys.unsetenv("NEONSTORE_HOME")
```

## Non-stacking files and low-level interface

At it's core, `neonstore` is simply a mechanism to download files from the NEON API.
While the `.csv` files from the Observation Systems (OS, e.g. bird count surveys),
and Instrument Systems (e.g. aquatic sensors) are typically stacked into large
tables, other products, such as the `.laz` and `.tif` images produced by the
airborne observation platform LIDAR and cameras may require a different approach.

```{r}
# Read in a large file list for illustration purposes
cper_data <- readr::read_csv("https://minio.thelio.carlboettiger.info/shared-data/neon_data_catalog.csv.gz")

## Typically one would read all files in local store, e.g. list.file(neon_dir())
df <- neon_filename_parser(cper_data$name)
```

```{r}
library(dplyr)
df %>% count(EXT, sort=TRUE)
```

We can take a look at all `laz` LIDAR files:

```{r}
df %>% 
  filter(EXT == "laz")
```

Note that many of the airborne observation platform (AOP) products, such as
these LIDAR files, do not include the PRNUM or REV components that make up part
of the `productCode`s used in the NEON `product` tables.

>>>>

The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@NAS_NEON; @NAS_GRAND].

@@

Provenance and performance are two central aspects to the design of `neonstore`.
A focus on provenance means that analyses can always be traced back to individual NEON data files, and supports both workflows that require access to the most recent or updated data and workflows that require repeated access to persistent, stable data files.
A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available working memory (RAM) on most machines: as many unified tables of even single products over all available sites and dates can range from 10 GBs into the terrabyte range. 
Because enabling analyses that could span the full spatial and temporal range of NEON's 81 sites over a 30 year horizon is a foundational motivation for NEON research, it essential that researchers be able to easily work with data across this full span. 
However, doing so requires techniques that go beyond workflows which expect data to be re-downloaded (provenance) or stored in memory for analysis (performance).

To address these challenges, `neonstore` implements a high-performance, provenance aware workflow by drawing on a host of key technologies, including cryptographic hashes, a Lightning Memory Mapped Database (LMDB) for metadata management, a columnar-oriented integrated relational database (duckdb) for data management, and optional support for remote access through parquet files cloud-native object stores to facilitate sharing large databases generated by `neonstore`. 
These technologies are seamlessly integrated behind a user-friendly R interface which requires no expertise in the underlying backend technology beyond a familiarity with R and the "tidy data" model used in the popular R packages, `dplyr` and `tidyr`.

@@

RStudio users can bring up an interactive list of available tables in the RStudio "Connections" pane using the function `neon_pane()` [Fig #1].

@@

```{r}
library(neonstore)
library(dplyr)
```

@@

One of the most important functions of `neonstore` is `neon_download()`, which downloads all files identified by a given product code, optionally constrained to specific site, date range, or file name pattern.  
For instance, we can download all of the NEON landbird survey data, identified as NEON data product `DP1.10003.001`: 
(The three-digit code identifies this as a Level 1 data product (DP1), product number 10003, revision 001, indicating this is the first and so far the only protocol used for bird survey sampling).

@@

Note that if we immediately attempt to re-download a subset of this data, `neonstore` detects that we have precisely matching files already downloaded to our local store:

@@

Note that the index also contains important provenance information, including the checksum (MD5 or CRC32), timestamp, and release tag indicated by NEON.
Internally, `neonstore` uses checksums to track the provenance of all files downloaded, which allows it to handle both the cases when NEON changes the name of a file but not the contents, and when NEON changes the contents of a file, even if the name remains unchanged. 
`neon_index()` takes optional arguments such as `product`, `start_date`, `end_date`, and `site` to allow for convenient sub-setting.

**Stacking datasets**

The highly atomized file structure breaks tables down by site and year/month as individual csv files (and other formats, including h5).  
Typical workflows will treat a collection of such files as a single data table, with columns indicating the site, year, and month of sampling.
When combining the collection of files into a single table (a process sometimes known as stacking), `neonstore` must handle several special cases:

1) Not all files have identical columns in identical order across all sites and months. This can break simple, high-performance file readers like `vroom` and `arrow`.
2) The data types of columns (string, numeric, date, timestamp, etc) must be inferred correctly. 
   This can be difficult when certain columns within individual files have no non-missing (`NA`) values.
3) Additional columns must be added to capture information (metadata) that is only available in the file name, such as siteID and position (for sensor-based products).
4) Most importantly, reading must be provenance-aware -- if there are two files with matching product, table, site, year-month and position (for sensor-based products) must resolve which file should be read in based on the timestamp provided, or the requested version.

`neonstore`'s stacking functions handle all of this for us.  When a version of the same data is detected with a revised timestamp compared to an earlier download, `neonstore` will notify us of the detection, but default to reading the most recent available version.

`neonstore` provides two alternative mechanisms for stacking data.
The simplest approach is `neon_read()`, which stacks all matching files into a single, in-memory `data.frame`.
Note that `neon_read()` can be given arguments to filter by site, date range, etc to read in a particular selection of data more quickly.
`neon_read()` always reads directly from the original, raw data files downloaded to the local neonstore.

```{r}
brd_countdata <- neon_read("brd_countdata-expanded")
```

For very large tables such as sensor data collections spanning many sites and years, users will require the increased performance provided by the high-performance, columnar-oriented relational database, `duckdb` [@duckdb].
To take advantage of this approach, we must first store the contents of the individual data files in the relational database using the function `neon_store()`:

```{r}
neon_store("brd_countdata")
```

This task needs to be done only once for each table or product after any new files have been downloaded.
Thanks to provenance tracking, any repeated calls to this function will not result in additional imports.
If a previously imported file has been deprecated by the publication of a matching file with newer timestamp,
`neon_store()` will remove the deprecated data and import the new data, with a warning saying so.

Once data has been imported into the local database, we can quickly load the table into R with `neon_table()`

```{r}
brd_countdata <- neon_table("brd_countdata")
```

This will be somewhat faster than `neon_read()`, since there is no need to parse the original data.
By default, this will still read all the data into a normal in-memory `data.frame` in R.
When working with very large data tables, it may be preferable or necessary to work with data
directly in duckdb, which can perform most `dplyr` and `tidyr` operations, such as `group_by()`, `summarise`, 
`filter`, and so forth, within the database.

**Working with very large data products**

```{r}
brd_countdata <- neon_table("brd_countdata", lazy = TRUE)
brd_countdata
```

Note that R displays what appears to be the same data frame as before, but with two essential differences.
R indicates that the displayed object is a `duckdb_connection`, and lists the number of rows only as `??`
(or within RStudio's rmarkdown interface, only the first 1000 rows will be displayed in the interactive table).
The number of rows is not displayed because this query is executed "lazily." Since the print method needs only
display the first 6 rows, the database does not execute the potentially expensive operation of counting all the rows
in the table, it merely displays the first 6. 
This approach is particularly pwoerful if we wish to first subset or summarise the data in a way that will result
in an overall smaller table.

Consider the example of a much larger table collected by sensor data; say, single-aspirated temperature collected every 30 mintues across all sites:
Note that we exploit pattern matching to provide a short-hand to the table name.
We download and import the 30m sampled table

```{r}
neon_download(product = "DP1.00002.001", table = "30") #Temp single aspirated
neon_store(product = "DP1.00002.001", table = "30") #Temp single aspirated
```

```{r}
temp <- neon_table(product = "DP1.00002.001",  table = "30", lazy=TRUE) #Temp single aspirated
```

```{r}
temp %>% 
  group_by(month) %>%
  summarise(mean = mean(),
            max = max(),
            min = min()) %>% 
  collect()
```

```{r}
index <- neon_index(product = "DP1.00002.001", table = "30") 
info <- index$path %>% fs::file_info()
n_files <- length(index$path)
n_bytes <- sum(info$size)
```

## Dependencies and optional dependencies

`neonstore`'s core dependencies are quite light, avoiding hard requirements on other packages which themselves bring in a large recursive list of dependencies. `DBI` [@DBI], `duckdb` [@duckdb], and `thor` [@thor] provide the core database interfaces; all have no recursive dependencies. `httr` [@httr] provides functionality required to interface with the NEON REST API, and `R.utils`, `zip`, and `vroom` [@vroom] provide parsers required to unpack the data. `progress` provides user-friendly progress bars and estimated run-times for slow functions.

Optional dependencies are required for specific tasks. `dplyr` is needed for lazy read in `neon_table()`.
`neonUtilities` [@neonUtilities] and `rhdf5` are required at this time for reading in eddy-covariance products.
`arrow` [@arrow] is required for remote database connections using `neon_remote()`.
`rstudioapi` is required to utilize `neon_pane()` to view the connections pane from within RStudio.
The remaining suggested dependencies are utilized within testing only.

## Appendix

Discuss the NEON filename convention parser, link to `neon_filename_parser.R` source for regex patterns.  
Discuss cases of NEON changing filename conventions.
